Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-07-02 doc/paper-lcr.tex
started re-write of theory

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For an isolated lens, the time delay
function can be calculated from calculated from 
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $\bmath{\theta}$ is the observed source position, $\bmath{\beta}$ is the 
unlensed source position, $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential, and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$
are, respectively, the angular diameter distances from observer to the lens, from observer
to the source, and from the lens to the source. The Fermat potential is given by
defined as
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right], 
\ee
where $\psi(\bmath{\theta}$ is the lens potential.
%The two term in Eq. \ref{eq:FP} comes from the geometric path difference as a result of
%the deflection, and the second term is the gravitational delay
%described by the lens potential $\psi(\bmath{\theta})$.

Strong lenses sit in our real Universe -- they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial allignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe. These additional structures induce an external shear
which can be recovered in the mass-modelling of the lens, but they can also introduce a convergence that
is undetectable from the image positions, shapes and relative fluxes. The mass-sheet degeneracy
\citep[see e.g.][for details]{Falco1985,Schneider2006}. This is a degeneracy in the strong lens
modeling corresponding to the addition of a mass sheet that contributes a convergence and
zero shear which, combined with a re-scaling of the original mass distribution, leaves the
predicted image positions and magnifications unchanged.
%Given a lens
%model $\kappa_{\rm lens}(\bmath{\theta})$, an alternative model described by the
%transformation 
%\be\label{eq:transform}
%\widetilde{\kappa}_{\rm lens}(\bmath{\theta}) = \lambda + (1-\lambda)\kappa_{\rm lens}(\bmath{\theta}),
%\ee
%where $\lambda$ is a constant, would fit the image positions, shapes and relative fluxes equally well. Physic%ally $\lambda$
%is the \flag{effective?} convergence of the mass-sheet and can take any real value. The mass-sheet degeneracy
%corresponds to a degeneracy between this external convergence ($\kappax$) and the mass
%normalization of the lens galaxy.
If the source brightness changes, the observed brightness in each image does not change
simultaneously - there is a time delay, $\Delta t$, between the brightness change in
each image. Despite the invariance of the image positions and fluxes, under the addition
of an external convergence the relative Fermat potential between the images is {\it not}
invariant, and so the time delay changes. As such it is necessary to include $\kappax$
in our lens modelling, if cosmological parameters are to be estimated accurately and
precisely \citep{suyu2010} from observed time delays.

\flag{How do kappas combine? this is what Sherry said in suyuetal2010 - To be specific, the prescription that we adopt for
  combining the effects of many mass sheets at redshifts $z_i$ with
  surface mass densities $\Sigma_i$ is $\kappax=\frac{4\pi
    G}{c^2}\displaystyle\sum_i\frac{\Sigma_i(D_i\vec\theta)D_i D_{i{\rm s}}}{D_{\rm s}}$.}

\be\label{eq:phitransform}
\Delta \widetilde{\phi_{\rm }}(\bmath{\theta},\widetilde{\bmath{\beta}_{\rm }}) 
= 
(1-\lambda) \Delta \phi_{\rm }
(\bmath{\theta},\bmath{\beta}_{\rm }).
\ee
 Thus if
there is any external convergence, $\kappax$, that is not included in the lens
modeling, then the value of $H_0$ --inferred assuming $\kappax = 0$ -- will be
$1-\kappax$ less than the true value of $H_0$:
\be 
\label{eq:MassSheet:H0bias}
H_0^{\rm{true}}=(1-\kappax) H_0^{{\kappax = 0}}.
\ee
This degeneracy is present because lensing observations only deliver
relative positions and fluxes. However we can overcome this degeneracy 
if we have knowledge of $\kappax$.

The lens potential, $\psi(\bmath{\theta})$, is related to the projected dimensionless
surface mass density, $\kappa(\bmath{\theta})$, and the physical projected surface mass density,
$\Sigma(D_{\rm d} \bmath{\theta})$, by 
\be \label{eq:psiKappaDiffRelan}
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}=\frac{1}{2}\nabla^2\psi(\bmath{\theta}),
\ee
where 
\be \label{eq:sigcrit} 
\Sigma_{\rm cr} \equiv \frac{c^2 D_{\rm s}}{4 \pi G D_{\rm d} D_{\rm ds}}.
\ee
Equation \ref{eq:Tsimp} can be rephrased as
\be
\Delta t(\bmath{\theta},\bmath{\beta})  =  \frac{\Dt}{c},
\ee 
where $\Dt \equiv (1+\zd) D_{\rm d} D_{\rm s}/D_{\rm ds}$ is the time-delay distance.
Angular diameter distances are given by
\begin{equation}
D_{\mathrm{ij}}~=~ {c/H_0 \over {(1+z_{\mathrm{j}})}}\left( {\mathrm{sinn}\!\left( \sqrt{|\Omega_k|} \int_{z_{\mathrm{i}}}^{z_{\mathrm{j}}} {\mathrm{d} z \over E(z)}\right)\over \sqrt{|\Omega_k|}}\right)
\label{eq:Da}
\end{equation}
where $\mathrm{sinn}(x)~=~\sin(x)$, $x$, or $\sinh(x)$ for open, flat, or closed universes respectively, and $E(z)$ is the normalised Hubble parameter:
\begin{subequations}\label{Eq:Ez}
\begin{align}
E(z) &\equiv {H(z) \over H_0}\\
E^{\Lambda{CDM}}&~=~\!\sqrt{{\Omega_{\mathrm{M}}(1\!+\!z)^{3}\!+\!\Omega_k(1\!+\!z)^2\!+\!\Omega_{\Lambda}}}
\end{align}
\end{subequations}
Since angular diameter distances are proportional to the inverse of the Hubble constant, $1/H_0$,
so is $\Dt$ proportional to $1/H_0$, hence a measurement of $\Dt$ can be used to infer $H_0$
and also the other parameters in Equations \ref{eq:Da} and \ref{eq:Ez}.

@@

In this work our aim is to quantify the lensing contribution from mass sheets that
are not physically associated with the lens galaxy -- mass-sheets lying infront or behind
the lens \citep{FassnachtEtal06}.

>>>>

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For an isolated lens, the time delay
function can be calculated from calculated from 
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $\bmath{\theta}$ is the observed source position, $\bmath{\beta}$ is the 
unlensed source position, $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential, and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$
are, respectively, the angular diameter distances from observer to the lens, from observer
to the source, and from the lens to the source. The Fermat potential is given by
defined as
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right], 
\ee
where $\psi(\bmath{\theta}$ is the lens potential, derived from the projected dimensionless
surface mass density, $\kappa(\bmath{\theta})$, and the physical projected surface mass density,
$\Sigma(D_{\rm d} \bmath{\theta})$, by 
\be \label{eq:psikappa}
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}=\frac{1}{2}\nabla^2\psi(\bmath{\theta}),
\ee
where 
\be \label{eq:sigcrit} 
\Sigma_{\rm cr} \equiv \frac{c^2 D_{\rm s}}{4 \pi G D_{\rm d} D_{\rm ds}}.
\ee

Strong lenses sit in our real Universe -- they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial allignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe. These additional structures induce an external shear
which can be recovered in the mass-modelling of the lens, but they can also introduce a convergence that
is undetectable from the image positions, shapes and relative fluxes; the mass sheet-degeneracy \citep[see e.g.][for details]{Falco1985,Schneider2006}. Similar to Eq. \ref{eq:psikappa}, the convergence from an isolated mass sheet is given by the projected surface mass density devided by the critical density, which is a function of redshift. 
\flag{but is this the kappa that enters the lens modelling?}
In the case of several mass sheets at redshifts $\{z_p\}$, with surface mass densities $\{\Sigma_p\}$, the effective convergence is given by the \flag{weighted?} sum :
\be \label{eq:kappasum}
\kappax=\frac{4\pi G}{c^2}\sum_p\frac{\Sigma_p(D_p\vec\theta)D_p D_{p{\rm s}}}{D_{\rm s}}.
\ee

If the source brightness changes, the observed brightness in each image does not change
simultaneously - there is a time delay, $\Delta t$, between the brightness change in
each image. Despite the invariance of the image positions and fluxes, under the addition
of an external convergence the relative Fermat potential between the images is {\it not}
invariant, and so the time delay changes. As such it is necessary to include $\kappax$
in our lens modelling, if cosmological parameters are to be estimated accurately and
precisely \citep{suyu2010} from observed time delays.

The time-delay distance is defined as $\Dt \equiv (1+\zd) D_{\rm d} D_{\rm s}/D_{\rm ds}$,
with which Equation \ref{eq:Tsimp} can be rephrased as
\be
\Delta t(\bmath{\theta},\bmath{\beta})  =  \frac{\Dt}{c},
\ee 
Since the time delays are proportional to $1/ (1+ \kappax)$, if there is any external convergence, $\kappax$, that is not included in the lens
modeling, then the time delay distance -- inferred assuming $\kappax = 0$ -- will be
$1-\kappax$ less than the true value of $\Dt$.
\be 
\label{eq:MassSheet:H0bias}
\Dt^{\rm{true}}=(1-\kappax) \Dt^{{\kappax = 0}}.
\ee
We can overcome this degeneracy if we have additional knowledge of $\kappax$.

@@

In this work our aim is to quantify the lensing contribution from mass sheets that
are not physically associated with the lens galaxy -- mass-sheets lying infront or behind
the lens. We also aim to quantify how uncertainties in the estimations of kappa
propogate into uncertainties on the cosmological parametres which can be inferred
from measurements of $\Dt$
